Stack Safety as a Security Property

### 2021-02-22 main.tex
Filled in first example

\section{Running Example}
\label{sec:running-example}

\begin{figure}[b]
\hfill
    \begin{minipage}{.3\textwidth}
\begin{verbatim}
main = {
  int x = 42, y = 0;
  y = f(0);
  print (x + y);
}
\end{verbatim}
    \end{minipage}
    \begin{minipage}{.3\textwidth}
\begin{verbatim}
f(z) = {
  int w = g(17);
  print z;
  return (z + w);
}
\end{verbatim}
    \end{minipage}
    \begin{minipage}{.3\textwidth}
\begin{verbatim}
g(v) = {
  return (v + 1);
}

\end{verbatim}
    \end{minipage}
\caption{A High-level Program With Nested Calls}
\label{fig:running-program}
\end{figure}

\newcommand*{\add}{\textsc{add}}
\newcommand*{\addi}{\textsc{addi}}
\newcommand*{\sw}{\textsc{sw}}
\newcommand*{\lw}{\textsc{lw}}
\newcommand*{\jal}{\textsc{jal}}
\newcommand*{\jalr}{\textsc{jalr}}
%\newcommand*{\rsp}{\textsc{sp}}
\newcommand*{\rra}{\textsc{ra}}
\newcommand*{\rout}{\textsc{out}}

\newcommand*{\tagInstr}{\textsc{instr}}
\newcommand*{\tagCall}{\textsc{call}}
\newcommand*{\tagHa}{\textsc{h1}}
\newcommand*{\tagHb}{\textsc{h2}}
\newcommand*{\tagRa}{\textsc{r1}}
\newcommand*{\tagRb}{\textsc{r2}}
\newcommand*{\tagRc}{\textsc{r3}}
\newcommand*{\tagNoDepth}{\textsc{unused}}
\newcommand*{\tagStackDepth}[1]{\textsc{stack} ~ #1}
\newcommand*{\tagPCDepth}[1]{\textsc{pc} ~ #1}
\newcommand*{\tagSP}{\textsc{sp}}

\newcounter{pcctr}
\setcounter{pcctr}{0}
%  \thepcctr & \stepcounter{pcctr} #1 & #2 & #3 & \text{#4}\\
\newcommand*{\row}[4]{
  \thepcctr & \stepcounter{pcctr} #2 & #3 & \text{#4}\\
}
\newcommand*{\tracerow}[5]{
  #1 & #2 & #3 & #4 & \text{#5}\\
}
\newcommand*{\summary}[6]{
\[
\begin{array}{cccccc}
  \PCname = #1 & r_0 = #2 & \rra = #3 & \rsp = #4 & r_4 = #5 & r_5 = #6
\end{array}
\]
}
\newcommand{\negate}{\textrm{-}}

\colorlet{lgray}{gray!40}

\newcommand{\instrc}{lgray}
\newcommand{\fprivc}{cyan}
\newcommand{\fcallc}{green}
\newcommand{\gprivc}{yellow}
\newcommand{\gcallc}{orange}
\newcommand{\hprivc}{red}
\newcommand{\unusedc}{gray}
\newcommand{\emptyoutc}{white} % or gray for consistency with unusedc ?
\newcommand{\fulloutc}{white}

\begin{figure}

%Starting Memory Layout
\begin{center}
\MemoryLabel{4em}{2em}{0}
\MemoryLabel{21em}{2em}{99}
\MemoryLabel{25em}{2em}{\SP}
\memory{3}{\instrc}[{\makebox[0pt]{Instructions}}]%
\hspace*{3pt}
$\cdots$
\memory{1}{\emptyoutc}[{\makebox[0pt]{Output}}]%
\memory{10}{\unusedc}[{\makebox[0pt]{Uninitialized}}]
~$\cdots$
\\
\end{center}
% Starting Register File
\summary{0}{0}{?}{100}{?}{?}
\vspace*{0.2em}
\ifaftersubmission\bcp{A very small nit: Change
  $\rsp-2$ to $\rsp\mathord{-}2$}\apt{??}\bcp{I meant remove spaces around +
and $-$ when then are used inside brackets.  I think it will look better.
But not critical.}\fi
\[
  \begin{array}{l|l|l|l}
    PC & \mathit{Instruction} & \mathit{Effect} & \\
    \hline
    \row{\addi ~ \rsp ~ \rsp ~ 2}{\rsp \leftarrow \rsp + 2}{\rsp \leftarrow 102}
        {Allocate private data for {\tt main}}
    \row{\addi ~ r_4 ~ r_0 ~ 42}{r_4 \leftarrow r_0 + 42}{r_4 \leftarrow 42}
        {Load 42 into $r_4$}
    \row{\sw ~ \rsp ~ r_4 ~ \negate 2}{[\rsp-2] \leftarrow r_4}{[100] \leftarrow 42}
        {Set {\tt x} to 42}
    \row{\sw ~ \rsp ~ r_0 ~ \negate 1}{[\rsp-1] \leftarrow r_0}{[101] \leftarrow 0}
        {Set {\tt y} to 0}
    \row{\addi ~ \rsp ~ \rsp ~ 2}{\rsp \leftarrow \rsp + 2}{\rsp \leftarrow 104}
        {Allocate {\tt f}'s call frame}
    \row{\sw ~ \rsp ~ r_0 ~ \negate 2}{[\rsp-2] \leftarrow r_0}{[102] \leftarrow 0}
        {Set argument to 0}
    \row{\sw ~ \rsp ~ r_0 ~ \negate 1}{[\rsp-1] \leftarrow r_0}{[103] \leftarrow 0}
        {Zero out result}
    \row{\jal ~ \rra ~ 20}{\rra \leftarrow \PCname + 1; ~ \PCname \leftarrow 20}{\rra \leftarrow \thepcctr; ~ \PCname \leftarrow 20}
        {Call ($\jal$ to) {\tt f}}
  \end{array}
  \]
  ~ \\
  ~\\
%  \apt{How's that description for last line?}
%  \apt{Where did these RISC-V insruction formats come from? I think some of them are just wrong, e.g.
%    $\jal$ takes a PC-relative argument, not an absolute one, and the proper way to code a return to $\rra$ is $\jalr  ~ \rra ~ r_0$.
%  But also the instruction operand ordering is awkward. Why not use standard AT\&T syntax?}
% At Call to G
% Memory Layout
\begin{center}
\MemoryLabel{4em}{2em}{0}
\MemoryLabel{21em}{2em}{99}
\MemoryLabel{43.5em}{2em}{\SP}
\memory{3}{\instrc}[{\makebox[0pt]{Instructions}}]%
\hspace*{3pt}
$\cdots$
\memory{1}{\emptyoutc}[{\makebox[0pt]{Output}}]%
\memory{2}{\fprivc}[{\makebox[0pt]{${\texttt{main}}_{\mathit{priv}}$}}]%
\memory{2}{\fcallc}[{\makebox[0pt]{{\tt f}'s call frame }}]%
\memory{6}{\unusedc}%[{\makebox[0pt]{Uninitialized}}]
~$\cdots$
\MemoryLabel{-46.5em}{0.75em}{42}
\MemoryLabel{-42.5em}{0.75em}{0}
\MemoryLabel{-38.5em}{0.75em}{0}
\MemoryLabel{-34.5em}{0.75em}{0}
\\
\end{center}
% Starting Register File
\summary{20}{0}{\thepcctr}{104}{42}{?}
%\vspace*{0.2em}
\setcounter{pcctr}{20}%
% G - Body - 1
\vspace*{0.2em}
  \[
  \begin{array}{l|l|l|l}
    \row{\addi ~ \rsp ~ \rsp ~ 2}{\rsp \leftarrow \rsp + 2}{\rsp \leftarrow 106}
        {Allocate stack space for {\tt w}}
    \row{\sw ~ \rsp ~ \rra ~ \negate 2}{[\rsp - 2] \leftarrow \rra}{[104] \leftarrow 8}
        {Save return address}
    % Zero out z?
    \row{\addi ~ \rsp ~ \rsp ~ 2}{\rsp \leftarrow \rsp + 2}{\rsp \leftarrow 108}
        {Allocate {\tt g}'s call frame}
    \row{\addi ~ r_4 ~ r_0 ~ 17}{r_4 \leftarrow r_0 + 17}{r_4 \leftarrow 17}
        {Store 17 into $r_4$}
    \row{\sw ~ \rsp ~ r_4 ~ \negate 2}{[\rsp - 2] \leftarrow r_4}{[106] \leftarrow 17}
        {Store 17 as the argument {\tt v}}
    \row{\sw ~ \rsp ~ r_0 ~ \negate 1}{[\rsp - 1] \leftarrow r_0}{[107] \leftarrow 0}
        {Zero out result}
    \row{\jal ~ \rra ~ 40}{\rra \leftarrow \PCname + 1 ; ~ \PCname \leftarrow 40}{\rra \leftarrow \thepcctr; ~ \PCname \leftarrow 40}
        {Call {\tt g}}
  \end{array}
  \]
  ~ \\
  ~ \\
% At Call to H
% Memory Layout
\begin{center}
\MemoryLabel{4em}{2em}{0}
\MemoryLabel{21em}{2em}{99}
\MemoryLabel{61.5em}{2em}{\SP}
\memory{3}{\instrc}[{\makebox[0pt]{Instructions}}]%
\hspace*{3pt}
$\cdots$
\memory{1}{\emptyoutc}[{\makebox[0pt]{Output}}]%
\memory{2}{\fprivc}[{\makebox[0pt]{${\texttt{main}}_{\mathit{priv}}$}}]%
\memory{2}{\fcallc}[{\makebox[0pt]{{\tt f}'s call frame }}]%
\memory{2}{\gprivc}[{\makebox[0pt]{${\texttt f}_{\mathit{priv}}$}}]%
\memory{2}{orange}[{\makebox[0pt]{{\tt g}'s call frame }}]%
\memory{2}{\unusedc}%[{\makebox[0pt]{Uninitialized}}]
~$\cdots$
\MemoryLabel{-46.5em}{0.75em}{42}
\MemoryLabel{-42.5em}{0.75em}{0}
\MemoryLabel{-38.5em}{0.75em}{0}
\MemoryLabel{-34.5em}{0.75em}{0}
\MemoryLabel{-30.5em}{0.75em}{8}
\MemoryLabel{-26.5em}{0.75em}{}
\MemoryLabel{-22.5em}{0.75em}{17}
\MemoryLabel{-18.5em}{0.75em}{0}
\\
\end{center}
% Starting Register File
\summary{40}{0}{\thepcctr}{108}{17}{?}
\setcounter{pcctr}{40}
\vspace*{0.2em}
% H - Body
\[
\begin{array}{l|l|l|l}
  \row{\addi ~ \rsp ~ \rsp ~ 1}{\rsp \leftarrow \rsp + 1}{\rsp \leftarrow 109}
      {Allocate space for return address}
  \row{\sw ~ \rsp ~ \rra ~ \negate 1}{[\rsp - 1] \leftarrow \rra}{[108] \leftarrow 27}
      {Save return address}
  \row{\lw ~ r_4 ~ \rsp ~ \negate 3}{r_4 \leftarrow [\rsp - 3]}{r_4 \leftarrow 17}
      {Load the argument ({\tt v}) into $r_4$}
  \row{\addi ~ r_4 ~ r_4 ~ 1}{r_4 \leftarrow r_4 + 1}{r_4 \leftarrow 18}
      {Increment $r_4$}\end{array}
\]
~\\
~\\
% At Middle of H
% Memory Layout
\begin{center}
\MemoryLabel{4em}{2em}{0}
\MemoryLabel{21em}{2em}{99}
\MemoryLabel{65.5em}{2em}{\SP}
\memory{3}{\instrc}[{\makebox[0pt]{Instructions}}]%
\hspace*{3pt}
$\cdots$
\memory{1}{\emptyoutc}[{\makebox[0pt]{Output}}]%
\memory{2}{\fprivc}[{\makebox[0pt]{${\texttt{main}}_{\mathit{priv}}$}}]%
\memory{2}{\fcallc}[{\makebox[0pt]{{\tt f}'s call frame }}]%
\memory{2}{\gprivc}[{\makebox[0pt]{${\texttt f}_{\mathit{priv}}$}}]%
\memory{2}{orange}[{\makebox[0pt]{{\tt g}'s call frame }}]%
\memory{1}{\hprivc}[{\makebox[0pt]{${\texttt g}_{\mathit{priv}}$}}]%
\memory{1}{\unusedc}%[{\makebox[0pt]{Uninitialized}}]
~$\cdots$
\MemoryLabel{-46.5em}{0.75em}{42}
\MemoryLabel{-42.5em}{0.75em}{0}
\MemoryLabel{-38.5em}{0.75em}{0}
\MemoryLabel{-34.5em}{0.75em}{0}
\MemoryLabel{-30.5em}{0.75em}{8}
\MemoryLabel{-26.5em}{0.75em}{}
\MemoryLabel{-22.5em}{0.75em}{17}
\MemoryLabel{-18.5em}{0.75em}{0}
\MemoryLabel{-14.5em}{0.75em}{27}
\\
\end{center}
% Register File
\summary{40\rb{44}}{0}{\thepcctr\rb{27}}{109}{18}{?}
\vspace*{0.2em}
\label{fig:running-trace-a}
\end{figure}

\begin{figure}
\[
\begin{array}{l|l|l|l}
  \row{\sw ~ \rsp ~ r_4 ~ \negate 2}{[\rsp - 2] \leftarrow r_4}{[107] \leftarrow 18}
      {Store $r_4$ as the return value}
  \row{\lw ~ \rra ~ \rsp ~ \negate 1}{\rra \leftarrow [\rsp - 1]}{\rra \leftarrow 27}
      {Load return address}
  \row{\addi ~ \rsp ~ \rsp ~ \negate 1}{\rsp \leftarrow \rsp - 1}{\rsp \leftarrow 108}
      {Deallocate {\tt g}'s local state}
  \row{\jalr ~ \rra ~ \rra ~ 0\apt{wrong}}{\PCname \leftarrow \rra}{\PCname \leftarrow 27}
      {Return to {\tt f} ($\jalr$)}
  \end{array}
\]

% After Return from H
% Memory Layout
\begin{center}
\MemoryLabel{4em}{2em}{0}
\MemoryLabel{21em}{2em}{99}
\MemoryLabel{61.5em}{2em}{\SP}
\memory{3}{\emptyoutc}[{\makebox[0pt]{Instructions}}]%
\hspace*{3pt}
$\cdots$
\memory{1}{\emptyoutc}[{\makebox[0pt]{Output}}]%
\memory{2}{\fprivc}[{\makebox[0pt]{${\texttt{main}}_{\mathit{priv}}$}}]%
\memory{2}{\fcallc}[{\makebox[0pt]{{\tt f}'s call frame }}]%
\memory{2}{\gprivc}[{\makebox[0pt]{${\texttt f}_{\mathit{priv}}$}}]%
\memory{2}{orange}[{\makebox[0pt]{{\tt g}'s call frame }}]%
\memory{2}{\unusedc}%[{\makebox[0pt]{Uninitialized}}]
~$\cdots$
\MemoryLabel{-46.5em}{0.75em}{42}
\MemoryLabel{-42.5em}{0.75em}{0}
\MemoryLabel{-38.5em}{0.75em}{0}
\MemoryLabel{-34.5em}{0.75em}{0}
\MemoryLabel{-30.5em}{0.75em}{8}
\MemoryLabel{-26.5em}{0.75em}{}
\MemoryLabel{-22.5em}{0.75em}{17}
\MemoryLabel{-18.5em}{0.75em}{18}
\\
\end{center}
% Starting Register File
\summary{27}{0}{\thepcctr\rb{27}}{108}{18}{?}
% \vspace*{0.2em}

\setcounter{pcctr}{27}
% H - Body
% \vspace*{0.2em}
\[
\begin{array}{l|l|l|l}
  \row{\lw ~ r_4 ~ \rsp ~ 0\rb{\negate 1}}{r_4 \leftarrow [\rsp \rb{- 1}]}{r_4 \leftarrow 18}
      {Load return value into $r_4$}
  \row{\sw ~ \rsp ~ r_4 ~ \negate 3}{[\rsp - 3] \leftarrow r_4}{[105] \leftarrow 18}
      {Set {\tt w} to return value}
  \row{\addi ~ \rsp ~ \rsp ~ \negate 2}{\rsp \leftarrow \rsp - 2}{\rsp \leftarrow 106}
      {Deallocate {\tt g}'s call frame}
  \row{\lw ~ r_5 ~ \rsp ~ \negate 4}{r_5 \leftarrow [\rsp - 4]}{r_5 \leftarrow 0}
      {Load argument {\tt z} to $r_5$}
  \row{\sw ~ \rout ~ r_5 ~ 0}{[\rout] \leftarrow r_5}{[99] \leftarrow 0}
      {Print {\tt z}}
  \row{\add ~ r_4 ~ r_4 ~ r_5}{r_4 \leftarrow r_4 + r_5}{r_4 \leftarrow 18}
      {Add $r_4$ and $r_5$}
  \row{\sw ~ \rsp ~ r_4 ~ \negate 3}{[\rsp - 3] \leftarrow r_4}{[103] \leftarrow 18}
      {Store $r_4$ as the return value}
  \row{\lw ~ \rra ~ \rsp ~ \negate 2}{\rra \leftarrow [\rsp - 2]}{\rra \leftarrow 8}
      {Load return address}
  \row{\addi ~ \rsp ~ \rsp ~ \negate 2}{\rsp \leftarrow \rsp - 2}{\rsp \leftarrow 104}
      {Deallocate {\tt f}'s local state}
  \row{\jalr ~ \rra ~ \rra ~ 0\apt{wrong}}{\PCname \leftarrow \rra}{\PCname \leftarrow 8}{Return to {\tt main} ($\jalr$)}
\end{array}
\]
% Return to main
% Memory Layout
\begin{center}
\MemoryLabel{4em}{2em}{0}
\MemoryLabel{21em}{2em}{99}
\MemoryLabel{43.5em}{2em}{\SP}
\memory{3}{\instrc}[{\makebox[0pt]{Instructions}}]%
\hspace*{3pt}
$\cdots$
\memory{1}{\fulloutc}[{\makebox[0pt]{Output}}]%
\memory{2}{\fprivc}[{\makebox[0pt]{${\texttt{main}}_{\mathit{priv}}$}}]%
\memory{2}{\fcallc}[{\makebox[0pt]{{\tt f}'s call frame }}]%
\memory{6}{\unusedc}%[{\makebox[0pt]{Uninitialized}}]
~$\cdots$
\MemoryLabel{-51em}{0.75em}{0}
\MemoryLabel{-47em}{0.75em}{42}
\MemoryLabel{-43em}{0.75em}{0}
\MemoryLabel{-39em}{0.75em}{0}
\MemoryLabel{-35em}{0.75em}{18}
\\
\end{center}
% Register File
\summary{8}{0}{\thepcctr\rb{8}}{104}{18}{0}
%\vspace*{0.2em}
\setcounter{pcctr}{8}
\[
\begin{array}{l|l|l|l}
  \row{\lw ~ \rsp ~ r_4 ~ \negate 1}{r_4 \leftarrow [\rsp - 1]}{r_4 \leftarrow 18}
      {Load return value to $r_4$}
  \row{\sw ~ \rsp ~ r_4 ~ \negate 3}{[\rsp - 3] \leftarrow r_4}{[101] \leftarrow 18}
      {Set {\tt y} to the result}
  \row{\addi ~ \rsp ~ \rsp ~ \negate 4}{\rsp \leftarrow \rsp - 2}{\rsp \leftarrow 102}
      {Deallocate {\tt f}'s call frame}
  \row{\lw ~ \rsp ~ r_5 ~ \negate 2}{r_5 \leftarrow [\rsp - 2]}{r_5 \leftarrow 42}
      {Load {\tt x} to $r_5$}
  \row{\add ~ r_4 ~ r_4 ~ r_5}{r_4 \leftarrow r_4 + r_5}{r_4 \leftarrow 60}
      {Add {\tt x} and {\tt y}}
  \row{\sw ~ \rout ~ r_4 ~ 0}{[\rout] \leftarrow r_4}{[99] \leftarrow 60}
      {Print {\tt x + y}}
\end{array}
\]
\vspace*{0.2em}
% Return to main
% Memory Layout
\begin{center}
\MemoryLabel{4em}{2em}{0}
\MemoryLabel{21em}{2em}{99}
\MemoryLabel{31.5em}{2em}{\SP}
\memory{3}{\instrc}[{\makebox[0pt]{Instructions}}]%
\hspace*{3pt}
$\cdots$
\memory{1}{\fulloutc}[{\makebox[0pt]{Output}}]%
\memory{2}{\fprivc}[{\makebox[0pt]{${\texttt{main}}_{\mathit{priv}}$}}]%
\memory{8}{\unusedc}%[{\makebox[0pt]{Uninitialized}}]
~$\cdots$
\MemoryLabel{-51em}{0.75em}{60}
\MemoryLabel{-47em}{0.75em}{60\rb{42}}
\MemoryLabel{-43em}{0.75em}{42\rb{18}}
\\
\end{center}
% Starting Register File
\summary{\thepcctr}{0}{37\rb{8}}{102}{60}{42}
\caption{Detailed Execution Trace
%  \rb{A suggestion for the description of $\jal$ and $\jalr$:
%    call(function label) [if labels are added next to the entrypoints]
%    or call(address of entrypoint) and return()?}
}
\label{fig:running-trace-b}
\end{figure}

\ifaftersubmission
\bcp{We really want readers to understand this trace.  I think we should
  spend a few more words walking through it in detail.}\leo{Not sure
  exactly what you mean here. It would take roughly a page to go through
  each instruction in detail - that's what the last column is for in my mind.
  Or do you have something else in mind?}
\bcp{I was thinking about just a few words, not a page.  But let's do this
  later.}
\fi

\Cref{fig:running-program} shows a high-level view of a running example,
written in C-like pseudo-code.  It consists of a {\tt main} function that
performs a call to some function {\tt f}, which in turn makes a nested call
to some function {\tt g} before returning, while performing some simple
arithmetic operations in the process. \Cref{fig:running-trace-b} shows an
execution trace of the corresponding machine instructions. These might have
been generated from the high-level code by a (rather simple-minded) compiler,
but it is only the behavior
of the machine instructions that we care about
here---\Cref{fig:running-trace-b} is our actual example.

To streamline the example, we use a small instruction set reminiscent of
RISC-V and make a few simplifying assumptions about the organization of
memory.
%
First, as noted above, since our goal is to formalize stack safety, we
ignore the heap
and assume that all variables are stack allocated.
%
We also assume that the machine is word- (not byte-) addressed, that program
instructions start from location 0, that the machine communicates with the
outside world through a single memory-mapped output port at address 99 (so
a {\tt print} statement will be compiled to a store to this
location), and that the stack begins at location 100 and grows upwards.
% \leo{Explain more?  How to connect to realistic one? Observations? DMA?
%   Interrupt handlers?}\bcp{Yes.}

A sketch of the initial memory layout appears at the top of
\Cref{fig:running-trace-b}: the instructions for {\tt main} begin at 0,
and the (uninitialized)
stack begins at 100. Directly below the memory sketch, we summarize the parts
of the register file that are relevant at this point in the example.
%
The rest of \Cref{fig:running-trace-b} traces the execution of the
program, showing the machine state at call and return points.
Instructions are intended to
be similar to RISC-V, but are described in a (hopefully more readable)
ad-hoc notation
rather than using assembler syntax. The $\jal$ (jump-and-link)
instruction used for calls atomically simultaneously the return address in a
register
and jumps to a specified new $\PCname$; the $\jalr$ (jump-and-link-via-register)
instruction used for returns jumps to the address stored in a register.
Register $r_0$ is hardwired to zero, so that, e.g., instruction 3 stores $0$
into address 101.

The code itself is straightforward, and should be largely self-explanatory, but
a few points deserve clarification. Functions return their result value on the
stack rather than in a register.  Stack frames are built in two parts:
space for local variables (which are all stack-allocated) and the return address is allocated and initialized at function entry, and deallocated at function exit; space for arguments and return values is allocated and initialized immediately prior to each function call and deallocated on return. The return value slot is always initialized
because some of our enforcement mechanisms will rely on this.
\ifaftersubmission\apt{this is bogus, but oh well}\fi
The return address is spilled to the stack even by leaf functions such as
{\tt g} that could just leave it in $\rra$.

%\apt{Need to clarify to what extent our properties depend on the
%particular compilation style, calling sequences, etc. illustrated here.}

>>>>

\section{Examples and Intuition}
\input{example1}
